Certifying, incremental type checking

### 2012-06-08 doc/papers/LFMTP2012/paper.tex
paper: abstract

\title{Incremental tactics}

\author{Matthias Puech\inst{1,2} \and Yann R\'egis-Gianas\inst{1}}
\institute{
  PPS, Team $\pi r^2$ (Universit\'e Paris Diderot, CNRS, INRIA)
  \and
  Department of Computer Science, University of Bologna
}

@@

\begin{abstract}
 %% 4 sentences:
 %% State the problem
 %% Say why it’s an interesting problem
 %% Say what your solution achieves
 %% Say what follows from your solution
\end{abstract}

@@

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \\&\qquad
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
system will generate this derivation, it will insert a coercion
$\cst{odd} \leq \cst{nat}$ at these two points. Taking advantage of
$\md_2$ and $\md_4$ the system succeeds with a new derivation made out
of these, without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

@@

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one into the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} x y {\s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

We exploit here an idea similar to run-time
\emph{contracts}: %TODO cite
there is no static guarantee that computations won't return ill-typed
values, but arguments and results are checked at run-time at the
boundaries of defined functions. This allows to omit formal
justifications, like the fact that we need to iterate twice the typing
of $P$ in $\recb M N x y P$, or that $B$ can't depend on $\md_x$ in
the $\lambda$ case. Of course this incurs on the performance; we will
see a compensation for it by the slicing mechanism described below in
\ref{sec:repr}.

@@

These typing rules use definitional equality judgement between two
families $\jea\mr\Gamma A {A'}$. This equivalence is usually defined
to be just $\alpha$-equivalence since objects are in
$\beta\eta$-canonical form. Here, objects can contain metavariables
defined in $\mr$, so we shall thread it. Moreover, we will see in the
next section how defined constants compute and thus influence
definitional equality. Definitional equality is defined later in Fig.
\ref{fig:def-eq}.

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects. We have the ability to construct
objects out of other, and deconstruct them by case analysis:
\begin{align*}
  T &\gequal
  \lam x T \gor
  U \\
  U &\gequal
  F \gor
  \matchin U \Gamma C \\
  C &\gequal \casebnobar P U \gor
  C\ \caseb P U \\
  P &\gequal
  \app H {\var x} \ldots\ {\var x}
\end{align*}

@@

\begin{figure*}
  \fbox{$\jcc\mr\Gamma T S A \mmr F$}
  \qquad
  {Computational cut}

  \begin{mathpar}
    \infer{
      \jm\mr\Gamma M {A_1} \mmr {M'}
      \and
      \jcc\mmr\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T {\msubst x {M'}}} S \mmmr F
    }{
      \jcc\mr\Gamma{\prd x {A_1} {A_2}}{\lam x T}{\spinecons M S}
      \mmmr F
    }

    \infer{
      \jcr\mr\Gamma U \mmr F
      \and
      \jm\mmr\Gamma F {\app {\cst a} S} \mmmr {F'}
    }{
      \jcc\mr\Gamma {\app {\cst a} S} U \spinenil \mmmr {F'}
    }
  \end{mathpar}

  \fbox{$\jcr\mr\Gamma U \mmr F$}
  \qquad
  {Computational reduction}

  \begin{mathpar}
    \infer{ }{
      \jcr\mr\Gamma{\app {\var x} S} \mr {\app {\var x} S}
    }

    \infer{
      \Sigma(c) = A; (\annsliceable\text{ or }\annnonsliceable)
    }{
      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
    }

    \infer{
      \Sigma(c) = A; \anndefined T
      \and
      \jcc\mr\Gamma T S A \mmr F
    }{
      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
    }

    \infer{
      \jcr\mr{\gconcat\Gamma{\Gamma'}} U \mmr {\app {H_i} {\spinecons
          {M_1}{\spinecons\ldots{M_{m_i}}}}}
      \and
      \jcr\mmr\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
          \msubst {x$_{m_i}$} {M_{m_i}}}} \mmmr F a
    }{
      \jcr\mr\Gamma{\matchin{U}{\Gamma'}{
          \casebnobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
                \ldots\ {\var x_{1m_1}}}}}{U_1}
          |\ \ldots\
          \caseb{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
              \ldots\ {\var x_{nm_n}}}}}{U_n}
        }}\mmr F
    }

@@

  \caption{Weak head-reduction}
  \label{fig:whr}
\end{figure*}

>>>>

\title{Certifying, incremental type checking}

\author{Matthias Puech\inst{1,2} \and Yann R\'egis-Gianas\inst{1}}
\institute{Univ Paris Diderot, Sorbonne Paris Cit\'e, PPS, UMR 7126
  CNRS, $\pi r^2$, INRIA Paris-Rocquencourt, F-75205 Paris, France
  \and Department of Computer Science, University of Bologna, 40126
  Bologna, Italy}

@@

\begin{abstract}
 %% 4 sentences:
 %% State the problem
  Type checkers are traditionally implemented as batch processes,
  taking a program and returning a binary information: well-typed or
  not.
 %% Say why it’s an interesting problem
  This laconicism presents a double problem: first, how can I trust
  that the checking algorithm (a complex program) is sound \wrt\ the
  declarative type system? Secondly, how can I avoid costly rechecks
  of the whole program when I only make a small, localized change?
 %% Say what your solution achieves
  A lightweight solution to both problems is letting the checker
  return a typing derivation: trust is then ensured by checking that
  this derivation is well-formed, and incrementality by the
  possibility of reusing already-computed pieces of derivations.
 %% Say what follows from your solution
  We present a system turning a type system and its checking algorithm
  into a safe and incremental one, consisting of a generic data
  structure to represent reusable derivations based on \LF\ and a
  general-purpose programming language to generate them.
\end{abstract}

@@

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} u v {\s{\var u}}}
  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function body $\s{\var u}$, and
he could reuse the pieces of derivation just built. Let us call these
respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
Provided there is an operator $\fget\md$ from the derivation to the
program it types, what really needs to be sent to the system is:
\begin{align*}
  &\finfer{}(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
  )
\ \text{,}
\end{align*}
that is, only the changed subterms of the program. Derivation $\md_2$
used to live in the empty context, we are then free to use it in the
new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
with the derivation $\finfer{\var u}$ inferred recursively for the new
recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
system generates this derivation, it inserts a coercion $\cst{odd}
\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
$\md_4$ the system succeeds with a new derivation made out of these,
without having to regenerate them because it knows that
$\finfer{\fget\md} = \md$.

@@

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\fget{\md_5}}
\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one into the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\fget{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} x y {\s{\var x}}}
  \\&\quad
  {(\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} z t {
      \app{\var{add}}\app{\var x} {\var z}})}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

We exploit here an idea similar to run-time \emph{contracts}
\cite{wadler2009well}: there is no static guarantee that computations
won't return ill-typed values, but arguments and results are checked
at run-time at the boundaries of defined functions. This allows to
omit formal justifications, like the fact that we need to iterate
twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
on $\md_x$ in the $\lambda$ case. Of course this incurs on the
performance; we will see a compensation for it by the slicing
mechanism described below in \ref{sec:repr}.

@@

These typing rules use definitional equality judgement between two
families $\jea\Delta\Gamma A {A'}$. This equivalence is usually defined
to be just $\alpha$-equivalence since objects are in
$\beta\eta$-canonical form. Here, objects can contain metavariables
defined in $\mr$, so we shall thread it. Moreover, we will see in the
next section how defined constants compute and thus influence
definitional equality. Definitional equality is defined later in Fig.
\ref{fig:def-eq}.

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects. We have the ability to construct
objects out of other, and deconstruct them by case analysis:
\begin{align*}
  T &\gequal
  \lam x T \gor
  U & \text{Term} \\
  U &\gequal
  F \gor
  \matchin U \Gamma C & \text{Atomic term}\\
  C &\gequal \casebnobar P U \gor
  C\ \caseb P U & \text{Branches}\\
  P &\gequal
  \app H {\var x} \ldots\ {\var x} &\text{Pattern}
\end{align*}

@@

\begin{figure*}
  \fbox{$\jcc\Delta\Gamma T S A {\Delta'} F$}
  \qquad
  {Term evaluation}

  \begin{mathpar}
    \infer[TLam]{
      \jm\Delta\Gamma M {A_1} {\Delta'} {M'}
      \and
      \jcc{\Delta'}\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T {\msubst x {M'}}} S {\Delta''} F
    }{
      \jcc\Delta\Gamma{\prd x {A_1} {A_2}}{\lam x T}{\spinecons M S}
      {\Delta''} F
    }

    \infer[TAtom]{
      \jcr\Delta\Gamma U {\Delta'} F
      \and
      \jm{\Delta'}\Gamma F {\app {\cst a} S} {\Delta''} {F'}
    }{
      \jcc\Delta\Gamma {\app {\cst a} S} U \spinenil {\Delta''} {F'}
    }
  \end{mathpar}

  \fbox{$\jcr\Delta\Gamma U {\Delta'} F$}
  \qquad
  {Atomic term evaluation}

  \begin{mathpar}
    \infer[UVar]{ }{
      \jcr\Delta\Gamma{\app {\var x} S} \Delta {\app {\var x} S}
    }

    \infer[UConst]{
      \Sigma(c) = A; (\annsliceable\text{ or }\annnonsliceable)
    }{
      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
    }

    \infer[UEval]{
      \Sigma(c) = A; \anndefined T
      \and
      \jcc\Delta\Gamma T S A {\Delta'} F
    }{
      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
    }

    \infer[UCase]{
      \jcr\Delta{\gconcat\Gamma{\Gamma'}} U {\Delta'} {\app {H_i} {\spinecons
          {M_1}{\spinecons\ldots{M_{m_i}}}}}
      \and
      \jcr{\Delta'}\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
          \msubst {x$_{m_i}$} {M_{m_i}}}} {\Delta''} F
    }{
      \jcr\Delta\Gamma{\matchin{U}{\Gamma'}{
          (\casebnobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
                \ldots\ {\var x_{1m_1}}}}}{U_1})\
          |\ \ldots\
          \caseb{(\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
              \ldots\ {\var x_{nm_n}}}}}{U_n})
        }}{\Delta''} F
    }

@@

  \caption{Evaluation of terms}
  \label{fig:whr}
\end{figure*}
